---
title: "ECON104 Assignment 2"
author: "Ahjin Kim, Jimin Kim, Bohyun Koo, Miguel Luis Martinez"
date: "May 12, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
CPI_data <- read.csv("C:/Users/15654/Documents/US CPI.csv")
```

## PART 1 - Time Series and Autocorrelation

1. Exploratory Data Analysis

\textcolor{blue}{(a) Briefly discuss the question you are trying to answer.}

The question we are trying to answer is, "How was inflation, as measured by the Consumer Price Index (CPI), evolved over time, and what model best captures its underlying dynamics for forecasting purposes?"

\textcolor{blue}{(b) Cite the dataset and give a summary of what the dataset is about; make sure it is a time-series.}

We got our dataset from Kaggle. The data used in this analysis is sourced from the U.S. Bureau of Labor Statistics, and the specific dataset is owned and managed by Arpit Verma. 
Verma, Arpit. (2021). *U.S. Inflation Data* [Data set]. https://www.kaggle.com/datasets/varpit94/us-inflation-data-updated-till-may-2021/data 


\textcolor{blue}{(c) First check for completeness and consistency of the data (if there are NAs or missing observations, replace with the value of the previous observation; make a note of this)}

```{r}
which(is.na(CPI_data))
```
The output of integer(0) signifies that the data is complete and there are no missing observations.

\textcolor{blue}{(d) Provide descriptive analyses of your variables. This should include the histogram with overlying density, boxplots, cross correlation. All figures/statistics must include comments.}

```{r histogram-density, echo=TRUE}
library(ggplot2)

ggplot(CPI_data, aes(x = CPI)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Histogram of CPI with Density Curve", x = "CPI", y = "Density")

```
The distribution of CPI values is right-skewed. This reflects the cumulative nature of inflation, with CPI values increasing over the years. The density curve peaks on the left, then tapers off toward the right, reinforcing the right-skewed nature of the data.


```{r}
ggplot(CPI_data, aes(x=CPI)) +
  geom_boxplot(fill="lightblue", color="darkgreen") +
  labs(title="Boxplot of CPI", x="CPI")

summary(CPI_data)
```
The boxplot shows that the median lies closer to the bottom of the interquartile range. The whiskers extend toward higher CPI values, but there are no outliers. 

The median is 33.10 while the mean is 82.64, which is much greater than the median, reinforcing that the data is right skewed. The range is 263.3, and the IQR is 130.3. We did not conduct a cross-correlation because our data analysis is univariate. 

2. Exploratory Data Analysis

\textcolor{blue}{(a) With tsdisplay or ggtsdisplay, for each variable, use its time series plot, ACF and PACF to comment on its stationarity (you can also decompose the time series; note if there is seasonality). To supplement this, use the appropriate Dickey-Fuller (unit root) test, to determine whether or not it is stationary. Note using its PACF what the suspected order might be.}

```{r}
library(tidyverse)
library(lubridate)
library(forecast)
library(tseries)
library(ggfortify)
library(ggpubr)

tsdisplay(cpi_ts, main="Original CPI Time Series")
```

There is a clear upward trend in CPI over time, which shows that the data is non-stationary because this would violate the stationarity assumption that mean is constant over time. The autocorrelation function (ACF) remains very high (close to 1) for all lags and decays slowly, which is a sign of non-stationarity. 

```{r}
adf.test(cpi_ts)
```
The Augmented Dickey-Fuller Test supports our conclusion of non-stationarity. The null hypothesis of this test is that the data is non-stationary. The p-value of 0.9881 is greater than the significant value of 0.05, so we fail to reject the null hypothesis and come to the conclusion that our data is not stationary. 



